Nonlinear Tidal Dissipation in Binary White Dwarfs

### 2019-02-12 proposal/FINESST/Su_FINESST_2019.tex
add convolve for amps

\def\Snospace~{\S{}} % hack to remove the space left after autorefs
\renewcommand*{\sectionautorefname}{\Snospace}
\renewcommand*{\appendixautorefname}{\Snospace}
\renewcommand*{\figureautorefname}{Fig.}
\renewcommand*{\equationautorefname}{Eq.}
\renewcommand*{\tableautorefname}{Tab.}

@@

Compact white dwarf (WD) binary systems, with orbital periods in the range of
minutes to hours, are important for a range of astrophysical problems. They are
the most important sources of gravitational waves (GWs) for the Laser
Interferometric Space Antenna (LISA)\cite{lisa}. They are also thought to
produce interesting optical transients such as tidal novae\cite{tidal_novae},
underluminous supernovae\cite{underlum}, and Ca-rich fast
transients\cite{carich}. Most importantly, they have been proposed as the likely
progenitors of type Ia supernovae (e.g.~\cite{Ia0,webbink} or more
recently\cite{Ia1,Ia2}). While presently only a few tens of compact WD binaries
are known\cite{lsst_wd}, \emph{Gaia} (currently gathering data) is expected to
expand the catalog to a few hundred\cite{lsst_wd} (results based on
\emph{Gaia}'s second data release have already begun to
appear\cite{gaiaDD,gaiaDD2}), and the Large Synoptic Survey Telescope (LSST,
first light scheduled for 2020) will likely detect a few thousand
more\cite{lsst_wd}. These observations will significantly advance the
understanding of WD binaries. My proposed theoretical and computational research
is well-timed to take advantage of these new advances.

In spite of the broad importance of WD binaries, the evolution of these systems
prior to their final merger is not well understood. Much of this uncertainty
comes from our imprecise understanding of tidal interactions, which play an
important role during a compact WD binary's inspiral\cite{fullerII}. Studies
show that these interactions manifest as tidal excitation of internal gravity
waves (IGW), waves in the WD fluid restored by the buoyancy force due to density
stratification\cite{fullerI}. As these waves propagate outwards towards the
surface, they grow in amplitude until they break, as do ocean waves on a shore,
and transfer both energy and angular momentum from the binary orbit to the outer
envelope of the WD\cite{fullerI,fullerII}.

Previous works have found that this tidal dissipation mechanism can generate
significantly more energy than thermal radiation from the WD surface alone and
is thus a major contributor to the WD energy budget\cite{fullerII,fullerIV}.
However, these works parameterized the wave breaking process without
hydrodynamical simulations. The details of tidal dissipation, namely the
location and spatial extent of the wave breaking, affect the observable outcome:
dissipation near the surface of the WD can be efficiently radiated away and
simply brightens the WD, while dissipation deep in the WD causes an energy
buildup that results in energetic flares\cite{tidal_novae}. Similar works in
other fields find that numerical simulations capturing the strongly nonlinear
wave breaking process find new behavior that cannot be described by linear and
weakly nonlinear theory\cite{winters1994,barker_ogilvie}. Such fully nonlinear
numerical simulations have not been performed for WDs.

\subsection{Proposed Research}

Characterizing the location and spatial extent of tidal dissipation in WD
binaries will require numerical simulation to capture the turbulent cascade to
small scales causing wave breaking. \textbf{I propose to study the dynamical
effects of tidal dissipation via nonlinear IGW breaking in binary WDs.} To
accomplish this, I aim to address the following three aims:
\begin{itemize}
    \item Characterize the location, spatial extent, and other properties of
        wave breaking in realistic WD models via direct numerical simulation.
        The location and spatial extent will furnish a simple yet effective
        parameterization of tidal dissipation. \autoref{s:2} describes the
        steps required for this aim.

    \item Predict signatures of tidal dissipation over a wide range of possible
        WD systems. In particular, I will describe the impact of tidal heating
        on the apparent temperature of a binary WD and maybe even the production
        of observable flares. \autoref{s:3} details my plan to do so.

@@

The current understanding of tidal synchronization in WD binaries is laid out
in~\cite{fullerII}: a tidally-excited train of IGW undergoes wave breaking in
the outer envelope of the WD, locally depositing angular momentum and
synchronizing the WD spin to the binary orbit. A similar process may also
operate in stellar binaries consisting of early type stars\cite{zahn75,gn89},
the only major difference being in the specifics of wave
excitation\footnote{While IGWs in stars are excited at their
radiative-convective boundaries, IGWs in WDs are excited at sharp changes in
composition.}. Nevertheless, direct numerical simulation of the wave breaking
process has not been performed in either of these systems. Since wave breaking
is a strongly nonlinear phenomenon, where a larger wave breaks down into many
smaller-scale waves, numerical simulation is paramount to an accurate
understanding of IGW breaking.

IGW breaking has been studied in detail in atmospheric sciences. The wave
breaking process proceeds as follows: initially, as the IGW reaches nonlinear
amplitudes, it breaks down via the parametric subharmonic instability and
transfers energy and angular momentum from the wave to the mean flow of the
fluid\cite{drazin}. Subsequently, after the mean flow velocity reaches the
horizontal phase velocity of the IGW, a critical layer through which the IGW
cannot propagate forms. A well-known calculation shows that the IGW is nearly
completely absorbed at this critical layer in the linear approximation and
endows the atmosphere with a mean horizontal flow\cite{booker_bretherton,hazel}.
However, when this mean flow absorption was numerically studied including full
nonlinear interactions, new phenomena not described by the linear theory
(reflection off the critical layer and sharpening of the mean flow) were
observed\cite{jones_num,winters1994} that affected the evolution of the
atmosphere over time. \emph{This highlights the importance of numerical
simulation in capturing the wave breaking process.}

To gain insight into the tidal dissipation process, I began by using the
spectral hydrodynamics code Dedalus\cite{dedalus} to study IGW breaking in a 2D
isothermal, stratified atmosphere. A spectral code like Dedalus is ideal for
simulating complex hydrodynamical phenomena, as spectral methods have no
inherent numerical viscosity and so better resolve the nonlinear cascade to
small length scales in wave breaking.

Working with Dr.\ Daniel Lecoanet (Princeton; one of the authors of the Dedalus
code) and my advisor, Prof.\ Dong Lai, I simulated an upward-propagating IGW
wavetrain excited at the bottom of the atmosphere. I observe the excited wave
breaking and depositing horizontal momentum in the fluid, causing the fluid to
acquire an average horizontal flow, consistent with previous
studies\cite{fullerII}. I derive simple formulae for the location and spatial
extent of the dissipation zone where the IGW is absorbed by the fluid. More
interestingly, I observe partial reflection of the IGW at the synchronization
layer\cite{me}, a phenomenon not considered in the current astrophysical
literature but consistent with the aforementioned results\cite{winters1994}. A
sample simulation is presented in \autoref{fig:nl_fluxes}. I am preparing these
results for publication\cite{me}.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.7\textwidth]{nl_fluxes.png}
    \caption{The evolution of the average horizontal flow velocity of the fluid
    $U_0$ (in units of $c_{ph,x}$ the horizontal phase velocity of the IGW) and
    horizontal momentum flux contained in the IGW $S_{px}(z)$ (in units of the
    total excited flux $S_0$) for one of my simulations. Different colored lines
    correspond to different times in the simulation. Times are measured in units
    of $N^{-1}$, the inverse of the Brunt-V\"ais\"al\"a frequency (or the
    buoyancy period), and heights in units of $H$ the scale height of the
    density stratification. IGW are excited at $z = 2H$ and propagate to higher
    $z$ before undergoing wave breaking. The sharp decrease in $S_{px}$ is
    the location of the critical layer where the IGW breaks and drives $U_0$
    towards $c_{ph, x}$; it moves to lower $z$ over time. The apparent decrease
    in $S_{px}$ at later times indicates reflection off the critical
    layer.}\label{fig:nl_fluxes}
\end{figure}

@@

It is therefore clear that full numerical modeling of IGW breaking is necessary
to fully characterize tidal dissipation. The first aim of my proposal is then:
\textbf{I will extend my preliminary results to characterize the dynamics of
nonlinear IGW breaking in realistic WDs.} Via numerical simulation, I will
develop simple models for tidal dissipation that can be used to study WD
evolution under tidal heating without performing further simulations. I will
continue my work in the following stages:
\begin{itemize}
    \item I will perform simulations examining the validity of my results in
        more realistic geometries, including polar and spherical geometries. I
        will continue to use Dedalus, which also supports these coordinate
        systems.

    \item I will extend my simulations to realistic WD models and equations of
        state such as those in~\cite{brassard1992}, continuing to track the
        location and spatial extent of the dissipation layer as well as any new
        phenomena. As WDs vary widely in composition and effective temperature,
        studying representative WD models is vital to obtaining a robust
        characterization of tidal dissipation. My selected WD models are those
        used in the current literature\cite{fullerII, fullerIV}.
\end{itemize}

>>>>

% \def\Snospace~{\S{}} % hack to remove the space left after autorefs
\renewcommand*{\sectionautorefname}{Section}
% \renewcommand*{\appendixautorefname}{\Snospace}
% \renewcommand*{\figureautorefname}{Fig.}
% \renewcommand*{\equationautorefname}{Eq.}
% \renewcommand*{\tableautorefname}{Tab.}

@@

Compact white dwarf (WD) binary systems, with orbital periods in the range of
minutes to hours, are important for a range of astrophysical problems. They are
the most important sources of gravitational waves (GWs) for the Laser
Interferometric Space Antenna (LISA)\cite{lisa}. They are also thought to
produce interesting optical transients such as underluminous
supernovae\cite{underlum}, Ca-rich fast transients\cite{carich}, and tidal
novae\cite{tidal_novae}. Most importantly, they have been proposed as the likely
progenitors of type Ia supernovae (e.g.~\cite{Ia0,webbink} or more
recently\cite{Ia1,Ia2}). While presently only a few tens of compact WD binaries
are known\cite{lsst_wd}, \emph{Gaia} (currently gathering data) is expected to
expand the catalog to a few hundreds\cite{lsst_wd} (results based on
\emph{Gaia}'s second data release have already begun to
appear\cite{gaiaDD,gaiaDD2}), and the Large Synoptic Survey Telescope (LSST,
first light scheduled for 2020) will likely detect a few thousand
more\cite{lsst_wd}. These observations will significantly advance the
understanding of WD binaries and their evolution. My proposed theoretical and
computational research is well-timed to take advantage of these new advances.

In spite of the broad importance of WD binaries, the evolution of these systems
prior to their final mergers is not well understood. Much of this uncertainty
comes from our imprecise understanding of tidal interactions, which play an
important role during a compact WD binary's inspiral\cite{fullerII}. Previous
studies have shown that these interactions manifest as tidal excitation of
internal gravity waves (IGW), waves in the WD fluid restored by the buoyancy
force due to density stratification\cite{fullerI}. As these waves propagate
outwards towards the WD surface, they grow in amplitude until they break, as do
ocean waves on a shore, and transfer both energy and angular momentum from the
binary orbit to the outer envelope of the WD\cite{fullerI,fullerII}.

Previous works have found that the dissipation of IGW can generate significantly
more energy than thermal radiation from the isolated WD surface and is thus a
major contributor to the WD energy budget\cite{fullerII,fullerIV}. However,
these works parameterized the wave breaking process in an ad hoc manner. The
details of dissipation, namely the location and spatial extent of the wave
breaking, affect the observable outcome: dissipation near the surface of the WD
can be efficiently radiated away and simply brightens the WD, while dissipation
deep in the WD envelope causes an energy buildup that results in energetic
flares\cite{tidal_novae}. Works in other fields based on numerical simulations
show that strongly nonlinear wave breaking process exhibits new behavior that
cannot be described by linear and weakly nonlinear
theory\cite{winters1994,barker_ogilvie}. Such fully nonlinear numerical
simulations have not been performed for WDs.

\subsection{Goals of Proposed Research}

Characterizing the location and spatial extent of tidal dissipation in WD
binaries will require numerical simulation to capture the turbulent cascade to
small scales that causes wave breaking. \textbf{I propose to study the dynamical
effects of tidal dissipation via nonlinear Internal Gravity Wave (IGW) breaking
in binary WDs.} There are three specific goals of my proposed research:
\begin{itemize}
    \item Characterize the location, spatial extent, and other properties of
        wave breaking in realistic WD models via direct numerical simulation.
        The location and spatial extent of wave breaking will furnish a simple
        yet effective parameterization of tidal dissipation in a range of WD
        models. \autoref{s:2} describes the steps required to acheive this goal.

    \item Predict signatures of tidal dissipation over a wide range of possible
        WD systems. In particular, I will study the impact of tidal heating
        on the luminosity of WDs in binaries and explore the possibility of
        producing of observable flares. \autoref{s:3} details my plan.

@@

The current understanding of tidal synchronization in WD binaries is laid out
in~\cite{fullerII}: tidal forces from the companion excites IGWs in the deep
envelope of the WDs. These IGW propagate outwards and undergo wave breaking in
the outer envelope of the WD, locally depositing angular momentum and
synchronizing the WD spin to the binary orbit. A similar process also
operates in binaries consisting of early type stars\cite{zahn75,gn89},
the only major difference being in the specifics of wave
excitation\footnote{While IGWs in massive stars are excited at their
radiative-convective boundaries, the excitation of IGWs in WDs is more gradual
and is associated with sharp composition changes in the stellar
envelope\cite{fullerII}.}. Nevertheless, direct numerical simulation of the wave
breaking process has not been performed in either of these systems. Since wave
breaking is a strongly nonlinear phenomenon, where a larger wave breaks down
into many smaller-scale waves, numerical simulation is paramount to an accurate
understanding of the tidal dissipation process.

IGW breaking has been studied in atmospheric sciences. The wave
breaking process proceeds as follows: Initially, as the IGW reaches nonlinear
amplitudes, it breaks down via the parametric subharmonic instability and
transfers energy and angular momentum from the wave to the mean flow of the
fluid\cite{drazin}; after the mean flow velocity reaches the
horizontal phase velocity of the IGW, a critical layer forms. Analytical
calculations show that the IGW is nearly completely absorbed at this critical
layer in the linear approximation and endows the atmosphere with a mean
horizontal flow\cite{booker_bretherton,hazel}. However, when this mean flow
absorption was numerically studied including full nonlinear interactions, new
phenomena not described by the linear theory (reflection off the critical layer
and sharpening of the mean flow) were observed\cite{jones_num,winters1994}; this
nonlinear behavior significantly affectedthe evolution of the atmosphere over
time. \emph{This highlights the importance of numerical simulation in capturing
the wave breaking process.}

To gain insight into the tidal dissipation process, I began by adapting the
spectral hydrodynamics code Dedalus\cite{dedalus} to study IGW breaking in a 2D
isothermal, stratified atmosphere. A spectral code like Dedalus is ideal for
simulating complex hydrodynamical phenomena, as spectral methods have no
inherent numerical viscosity and so can more accurately resolve the nonlinear
cascade to small length scales in wave breaking.

Working with Dr.\ Daniel Lecoanet (a post-doc at Princeton University, and one
of the authors of the Dedalus code) and my advisor, Prof.\ Dong Lai, I have
simulated the nonlinear evolution of an upward-propagating IGW wavetrain excited
at the bottom of the atmosphere. My simulations show the waves breaking and
depositing horizontal momentum in the fluid, causing the fluid to acquire an
average horizontal flow, consistent with previous studies\cite{fullerII}. I have
derived simple formulae for the location and spatial extent of the dissipation
zone where the IGW is absorbed by the fluid. More interestingly, my simulation
reveals a partial reflection of the IGW at the critical layer\cite{me}, a
phenomenon not considered in the current astrophysical literature but consistent
with the aforementioned results\cite{winters1994}. A sample simulation is
presented in \autoref{fig:nl_fluxes}. I am preparing these results for
publication\cite{me}.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{nl_fluxes.png}
    \caption{The evolution of the average horizontal flow velocity of the fluid
    $U_0$ (in units of $c_{ph,x}$ the horizontal phase velocity of the IGW) and
    horizontal momentum flux contained in the IGW $S_{px}(z)$ (in units of the
    total excited flux $S_0$) for one of my simulations. Different lines
    correspond to different times in the simulation. Times are measured in units
    of $N^{-1}$, the inverse of the Brunt-V\"ais\"al\"a frequency (or the
    buoyancy period), and heights in units of $H$ (the scale height of the
    density stratification). IGWs are excited at $z = 2H$ and propagate to
    higher $z$ before undergoing wave breaking. The sharp decrease in $S_{px}$
    is the location of the critical layer where the IGW breaks and drives $U_0$
    towards $c_{ph, x}$; it moves to lower $z$ over time. The apparent decrease
    in $S_{px}$ at later times indicates wave reflection off the critical
    layer.}\label{fig:nl_fluxes}
\end{figure}

@@

It is clear that full numerical modeling of IGW breaking is necessary to fully
characterize tidal dissipation. The first aim of my proposal is to
\textbf{extend my preliminary results concerning IGW breaking in stratified
atmospheres to characterize the dynamics of nonlinear IGW breaking in realistic
WD models.} Via numerical simulation, I will develop models for tidal
dissipation that can be used to study long-term WD evolution under tidal heating
without expensive hydrodynamical simulations. I will continue my work in the
following stages:
\begin{itemize}
    \item I will perform simulations examining the validity of my results in
        spherical geometries to capture tidal effects in WD binaries. I will
        continue to use Dedalus, which supports spherical coordinates. Although
        3D simulations are necessarily more complex than my 3D simulations of
        stratified atmospheres, the underlying dynamics of IGW breaking are the
        same.

    \item I will extend my simulations to realistic WD models and equations of
        state such as those in~\cite{brassard1992} as well as those generated by
        MESA (see \autoref{s:3} below), continuing to track the location and
        spatial extent of the dissipation layer as well as any new phenomena. As
        WDs vary widely in composition and effective temperature, studying
        representative WD models is vital to obtaining a robust characterization
        of tidal dissipation.
\end{itemize}
